---
output: 
  github_document:
    toc: true
    toc_depth: 4
    fig_width: 8
    fig_height: 4
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.retina = 2,
  fig.align = "center",
  dpi = 300,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  cache = FALSE,
  dev = "png"
)
```

# Milk Quality Spectroscopy {.unnumbered}

<!-- badges: start -->

<!-- badges: end -->

```{r}
suppressPackageStartupMessages({
  library(HotellingEllipse)
  library(ConfidenceEllipse)
  library(tidyverse)
  library(magrittr)
  library(reticulate)
  library(viridis)
  library(arrow)
  library(tidymodels)
  library(future)
  library(corrplot)
  library(vip)
  library(patchwork)
  library(yardstick)
  library(skimr)
  library(GGally)
  library(FactoMineR)
  library(factoextra)
  library(gridExtra)
  library(RColorBrewer)
  library(knitr)
  library(naniar)
  library(janitor)
  library(ggforce)
  library(ranger)
  library(glmnet)
  library(xgboost)
  library(mixOmics)
})
```

```{r}
tidymodels_prefer()
theme_set(theme_light(base_size = 12))
```

# Dataset

The data is structured by sensor/stall with each sensor directory containing:

-   **Lab results** (CSV format with tube_no as primary key)
-   **Sensor configuration** (wavelengths, calibration coefficients)
-   **Spectral measurements** (Parquet files, one per milk sample)
-   **Dark spectra** (reference measurements)

## Key data components

**Lab Results**: Milk analysis including fat percentage, protein, somatic cell count and, lactose linked to specific cows and milking sessions.

**Spectral Data**: Raw 16-bit spectral arrays captured during milking, with metadata like temperature, LED current, and integration time. Each spectrum is classified as "dark," "sample," or "empty."

**Sensor Info**: Each sensor has unique wavelength calibrations and measurement parameters that aren't standardized across sensors.

## Important limitations

-   Temperature and LED measurements are raw ADC values, not standardized between sensors
-   Each sensor measures different wavelengths
-   Some timing discrepancies remain due to clock source differences

## Data evolution

The dataset has evolved from initial CSV format to Parquet compression, with spectral data consolidated into array columns rather than individual wavelength columns for more efficient storage and processing.

# Data processing

```{r}
set.seed(123)
```

```{r}
dataset_path <- "Dataset/"
functions_path <- "~/Documents/GitHub/milk-quality-spectroscopy/R/"
```

```{r}
source(paste0(functions_path, "read_sensor_csvs.R"))
source(paste0(functions_path, "read_sensor_parquets.R"))
source(paste0(functions_path, "subtract_dark_spectrum.R"))
source(paste0(functions_path, "process_nir_data.R"))
```

```{r message=FALSE, warning=FALSE}
nir_data <- process_nir_data(verbose = FALSE)
```

```{r}
sensor_wavelengths_list <- nir_data$sensor_data %>%
  select(sensor, wavelengths_nm) %>%
  mutate(
    wavelength_chars = 
      str_remove_all(wavelengths_nm, "\\[|\\]") %>%
      str_split(",\\s*")
    ) %>%
  select(sensor, wavelength_chars) %>%
  deframe()
```

```{r}
sensor_wavelengths_list %>% glimpse()
```

```{r}
spec_data <- nir_data$corrected_spectra %>%
  filter(spec_type == "SAMPLE") %>%
  select(tube_number, sensor, spec_array) %>%
  modify_at("tube_number", as_factor) %>%
  modify_at("sensor", as_factor) %>%
  mutate(
    spec_valid = case_when(
      map_lgl(spec_array, is.null) ~ FALSE,
      map_int(spec_array, length) == 0 ~ FALSE,
      map_lgl(spec_array, ~ all(is.na(.x))) ~ FALSE,
      TRUE ~ TRUE
    )
  ) %>%
  filter(spec_valid) %>%
  rowwise() %>%
  mutate(
    df = list({
      tryCatch(
        {
          values <- as.numeric(unlist(spec_array))
          valid_values <- values[!is.na(values)]
          if (length(valid_values) > 0) {
            setNames(valid_values, paste0("X", seq_along(valid_values)))
          } else {
            NULL
          }
        },
        error = function(e) {
          NULL
        }
      )
    })
  ) %>%
  ungroup() %>%
  filter(!map_lgl(df, is.null)) %>%
  select(-spec_array, -spec_valid) %>%
  unnest_wider(df) %>%
  relocate(sensor, .before = tube_number)
```

```{r}
spec_data
```

```{r}
lab_data <- nir_data$lab_results %>%
  select(
    sensor,
    tube_number,
    fat_percent,
    protein_percent,
    scc_thous_per_ml,
    lactose_percent,
    milk_wt_lbs,
    other_solids_percent,
    solids_not_fat_percent,
    total_solids_percent
  ) %>%
  modify_at("tube_number", as_factor) %>%
  modify_at("sensor", as_factor) %>%
  filter(
    !is.na(fat_percent),
    !is.na(protein_percent),
    !is.na(scc_thous_per_ml),
    !is.na(lactose_percent)
  ) %>%
  rename(
    fat = fat_percent,
    protein = protein_percent,
    scc = scc_thous_per_ml,
    lactose = lactose_percent,
    weight = milk_wt_lbs,
    other_solids = other_solids_percent,
    nonfat_solids = solids_not_fat_percent,
    tot_solids = total_solids_percent
  )
```

```{r}
lab_data
```

```{r}
targets <- lab_data %>% select(fat, protein, scc, lactose) %>% names()
meta <- lab_data %>% select(-sensor, -tube_number, -all_of(targets)) %>% names()
```

```{r}
modeling_data <- spec_data %>%
  left_join(lab_data, by = c("sensor", "tube_number")) %>%
  filter(if_all(starts_with("X_"), ~ !is.na(.x))) %>%
  relocate(sensor, .before = tube_number) %>%
  relocate(all_of(targets), .after = tube_number) %>%
  relocate(all_of(meta), .after = all_of(targets))
```

```{r}
cat("Final dataset dimensions:", nrow(modeling_data), "x", ncol(modeling_data), "\n")
cat("Number of wavelenghts:", sum(grepl("^X", names(modeling_data))), "\n")
```

## Observations by Sensor

```{r}
sensor_summary <- modeling_data %>%
  select(sensor, tube_number, starts_with("X")) %>%
  group_by(sensor, tube_number) %>%
  summarise(
    observations = n(),
    variables = ncol(.) - 2,
    total_cells = n() * ncol(.),
    missing_values = sum(is.na(.)),
    completeness = round((1 - sum(is.na(.)) / (n() * ncol(.))) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(sensor)
```

```{r}
sensor_summary
```

```{r fig.height=5}
sensor_summary %>%
  group_by(sensor) %>%
  summarise(total_obs = sum(observations), .groups = "drop") %>%
  mutate(sensor_num = as.numeric(str_extract(sensor, "\\d+"))) %>%
  arrange(sensor_num) %>%
  ggplot(aes(x = reorder(sensor, sensor_num), y = total_obs, fill = sensor)) +
  geom_col(alpha = 0.8) +
  geom_text(
    aes(label = scales::comma(total_obs)),
    vjust = -0.3,
    size = 3.5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = colorRampPalette(brewer.pal(9, "Blues"))(12)) +
  scale_x_discrete(
    limits = paste0("sensor_", 1:12),
    labels = paste("Sensor", 1:12)
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Total Observations by Sensor",
    subtitle = "Comparison across all 12 sensors",
    x = "",
    y = "Total Observations"
  ) +
  theme_grey() +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```

```{r message=FALSE, warning=FALSE}
sensor_summary %>%
  mutate(sensor_num = as.numeric(str_extract(sensor, "\\d+"))) %>%
  group_by(sensor) %>%
  arrange(tube_number) %>%
  mutate(
    tube_position = row_number(),
    max_tubes = max(tube_position)
  ) %>%
  ungroup() %>%
  ggplot(
    aes(
      x = tube_position, 
      y = reorder(sensor, -sensor_num), 
      fill = observations
      )
    ) +
  geom_tile(color = "white", size = 0.3) +
  scale_fill_gradient2(
    low = "lightblue", mid = "yellow", high = "red",
    midpoint = median(sensor_summary$observations),
    name = "Observations"
  ) +
  scale_y_discrete(
    limits = paste0("sensor_", 12:1),
    labels = paste("Sensor", 12:1)
  ) +
  labs(
    title = "Observations Heatmap: All Sensors",
    subtitle = "X-axis shows tube position (1st, 2nd, 3rd, etc.) within each sensor",
    x = "Tube Position (ordered within sensor)",
    y = ""
  ) +
  theme_light() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 8)
  ) +
  scale_x_continuous(
    breaks = seq(
      1, 
      max(
        sensor_summary %>% 
          group_by(sensor) %>% 
          summarise(n = n()) %>% 
          pull(n)
        ), 
      by = 5
      )
    )

```

## Measurement timeline

```{r}
heatmap_data <- nir_data$corrected_spectra %>%
  mutate(datetime = as.POSIXct(datetime)) %>%
  filter(!is.na(datetime)) %>%
  filter(spec_type == "SAMPLE") %>%
  drop_na() %>%
  mutate(
    date_hour = floor_date(datetime, "hour"),
    sensor_num = as.integer(str_extract(sensor, "\\d+"))
  ) %>%
  count(date_hour, sensor_num, name = "n_measurements")

all_sensors <- as.integer(1:12)
hour_range <- seq(
  floor_date(min(heatmap_data$date_hour), "hour"),
  ceiling_date(max(heatmap_data$date_hour), "hour") - hours(1),
  by = "hour"
)

measurement_data <- nir_data$corrected_spectra %>%
  mutate(datetime = as.POSIXct(datetime)) %>%
  filter(!is.na(datetime)) %>%
  filter(spec_type == "SAMPLE") %>%
  drop_na() %>%
  arrange(datetime, sensor)
```

```{r}
sequence_plot <- measurement_data %>%
  ggplot(aes(x = datetime, y = factor(sensor), color = factor(sensor))) +
  geom_linerange(
    aes(
      ymin = as.numeric(factor(sensor)) - 0.4,
      ymax = as.numeric(factor(sensor)) + 0.4
    ),
    linewidth = 0.1,
    alpha = 0.8
  ) +
  ggsci::scale_color_d3(palette = "category20") +
  scale_y_discrete(
    limits = paste0("sensor_", 12:1),
    labels = paste("Sensor", 12:1)
  ) +
  scale_x_datetime(
    date_labels = "%m/%d\n%H:%M",
    date_breaks = "6 hours",
    minor_breaks = "1 hour",
    expand = c(0.01, 0)
  ) +
  labs(
    title = "Sensor Sequence Temporal Pattern: Measurement Timeline",
    subtitle = sprintf(
      "Total measurements: %d | Period: %s to %s",
      nrow(measurement_data),
      format(min(measurement_data$datetime), "%Y-%m-%d %H:%M"),
      format(max(measurement_data$datetime), "%Y-%m-%d %H:%M")
    ),
    x = "Date/Time",
    y = "",
    caption = "Each vertical line = one measurement | Clear gaps show missing sensors in sequence\nLook for vertical alignment (simultaneous) vs diagonal patterns (sequential)"
  ) +
  theme_light() +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "gray40"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_text(size = 9),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

```{r}
print(sequence_plot)
```

## NIR Spectra

```{r}
source(paste0(functions_path, "visualize_nirs_time_series.R"))
```

```{r message=FALSE, warning=FALSE}
visualize_nirs_time_series(nir_data, "sensor_1", min_measurements = 2, seed = 100)
```

```{r}
wavelengths_sorted <- sort(as.numeric(sensor_wavelengths_list$sensor_1))
```

```{r}
modeling_data %>%
  filter(sensor == "sensor_1") %>%
  select(tube_number, fat, starts_with("X")) %>%
  slice_sample(n = 30) %>%  # Only take 5 samples
  setNames(c("tube_number", "fat", wavelengths_sorted)) %>%
  pivot_longer(
    cols = !c(fat, tube_number),
    names_to = "Wavelength",
    values_to = "Intensity"
  ) %>%
  mutate(Wavelength = as.numeric(Wavelength)) %>%
  ggplot(aes(x = Wavelength, y = Intensity, group = tube_number, color = fat)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c(name = "Fat Content", option = "plasma") +
  labs(
    title = "NIRS Spectra by Fat Content",
    x = "Wavelength (nm)",
    y = "Intensity"
  ) +
  theme_minimal()
```

# Exploratory Data Analysis

## Descriptive statistics for milk quality target variables

```{r}
lab_data %>% 
  select(all_of(targets)) %>%
  specProc::summaryStats()
```

```{r}
cor_matrix <- lab_data %>%
  select(-sensor, -tube_number) %>%
  cor(use = "complete.obs")

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  order = "hclust",
  tl.cex = 0.8,
  tl.col = "black",
  addCoef.col = "black",
  number.cex = 0.8,
  title = "Correlation matrix of all variables",
  mar = c(0, 0, 2, 0)
)
```

```{r message=FALSE, warning=FALSE, fig.height=10, fig.width=10}
lab_data %>%
  select(-tube_number) %>%
    ggpairs(
      columns = 1:5,
      ggplot2::aes(colour = sensor),
      title = "Pairwise relationships between target variables"
    )
```

```{r message=FALSE, warning=FALSE}
p <- lab_data %>% 
  select(all_of(targets)) %>%
  specProc::adjusted_boxplot()

p + facet_wrap(~ variable, scales = "free", ncol = 4) +
  scale_fill_viridis_d(option = "viridis") +
  labs(title = "Overall distribution of milk components") +
  theme_light() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(size = 10, color = "black", face = "bold"),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
    )
```

Counting the number of zero values across all target variables:

```{r}
lab_data %>% 
  select(all_of(targets)) %>%
  summarise(across(everything(), ~ sum(.x == 0, na.rm = TRUE)))
```

These zero values probably represent missing or invalid data rather than true biological zeros, and might be better handled through imputation or exclusion rather than including them in the modeling.

```{r message=FALSE, warning=FALSE}
lab_data %>% 
  select(all_of(targets)) %>%
  # filter(if_all(everything(), ~ .x > 0)) %>% 
  specProc::outlierplot(show.mahal = TRUE) +
  theme(strip.text = element_text(size = 10, color = "black", face = "bold"))
```

The color gradient shows the Mahalanobis distance, while the triangular shape indicates multivariate outliers, i.e. samples that are unusual when considering all targets together. Most variables show tight clustering. Fat, lactose, other_solids, and protein all show the majority of data points clustered tightly around zero with relatively few outliers. SCC shows much more variability, showing a broader spread of Mahalanobis distances.

The yellow triangles in the fat, lactose, other_solids, and protein panels, represent extreme outliers with very high Mahalanobis distances. They have unusually low fat, lactose, other solids, and protein content compared to the normal population. Moreover, the fact that fat, protein, and lactose all have exactly 11 zeros suggests these might be the same samples with missing or problematic measurements. The 11 samples with zeros across multiple variables would definitely appear as extreme multivariate outliers (explaining some of those yellow triangular points at very negative values).

SCC has a highly skewed distribution based on its box plot. This extreme right skew in SCC explains why it showed such different patterns in the outlier detection plot. The other variables (fat, protein, lactose) show more symmetric, normal-like distributions in their box plots, which is why their outliers in the Mahalanobis plot were primarily samples with unusually low values. For SCC, however, the natural distribution is so skewed that most "normal" samples cluster near zero. The outliers detected by the robust Mahalanobis method are likely the samples with moderately high SCC values that, while not extreme on the raw scale, are unusual in the context of the multivariate relationship with other milk components.

```{r}
transf_list <- lab_data %>%
  select(all_of(targets)) %>%
  specProc::robustBCYJ(var = c("scc", "lactose"))
```

```{r message=FALSE, warning=FALSE}
transf_list %>%
  pluck("transformation") %>%
  bind_cols(lab_data %>% select(fat, protein)) %>%
  relocate(c("scc", "lactose"), .after = protein) %>%
  specProc::adjusted_boxplot() +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  scale_fill_viridis_d(option = "viridis") +
  labs(title = "Robust Box-Cox transformation of SCC and lactose") +
  theme_light() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(size = 10, color = "black", face = "bold"),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
    )
```

```{r message=FALSE, warning=FALSE}
transf_list %>%
  pluck("transformation") %>%
  bind_cols(lab_data %>% select(fat, protein)) %>%
  relocate(c("scc", "lactose"), .after = protein) %>%
  specProc::outlierplot(show.outlier = FALSE, show.mahal = TRUE) +
  theme(strip.text = element_text(size = 10, color = "black", face = "bold"))
```

After the Box-Cox transformation, SCC now shows a more symmetric distribution around zero, rather than the extreme right skew we saw in the original data. The transformation has successfully normalized the SCC distribution. The dark points (high Mahalanobis distance outliers) appear across all panels, suggesting these represent samples with genuinely unusual multivariate patterns rather than just extreme values driven by SCC's original skewed distribution.

## PCA analysis of milk quality targets

A principal component analysis (PCA) was conducted on the four key milk composition variables. The analysis aimed to understand the underlying structure of relationships between these variables and assess potential multicollinearity for subsequent modeling efforts.

```{r}
pca_model <- lab_data %>%
  select(all_of(targets), other_solids) %>%
  PCA(scale.unit = TRUE, graph = FALSE)
```

```{r}
eigenvalues <- get_eigenvalue(pca_model)
```

```{r}
p1 <- pca_model %>%
  fviz_eig(
    addlabels = TRUE,
    ylim = c(0, 60),
    title = "Scree Plot",
    xlab = "Principal Components",
    ylab = "Percentage of Variance Explained"
  ) +
  theme_light() +
  labs(caption = "Red line indicates 10% variance threshold")

p2 <- eigenvalues %>%
  as.data.frame() %>%
  mutate(component = 1:nrow(.)) %>%
  ggplot(aes(x = component, y = cumulative.variance.percent)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  scale_x_continuous(breaks = 1:nrow(eigenvalues)) +
  labs(
    title = "Cumulative Variance Explained",
    x = "Principal Components",
    y = "Cumulative Variance (%)",
    caption = "Red line indicates 80% variance threshold"
  ) +
  theme_light()
```

```{r}
p1 | p2
```

The PCA extracted two principal components that collectively explained 80.5% of the total variance in the dataset. The first principal component accounted for 58.9% of the variance, while the second component explained an additional 21.6%.

```{r}
pca_contrib <- function(model, axes, title, fill_color, top = 10) {
  p <- model %>%
    fviz_contrib(
      choice = "var",
      axes = axes,
      top = top,
      title = title,
      fill = fill_color,
      color = fill_color
    ) +
    labs(x = "") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  return(p)
}
```

```{r}
pca_contrib(pca_model, axes = 1, title = "Dim1", fill_color = "steelblue") |
  pca_contrib(pca_model, axes = 2, title = "Dim2", fill_color = "coral") |
  pca_contrib(pca_model, axes = 1:2, title = "Dim1 and Dim2", fill_color = "darkgreen")
```

Dim1 represents a milk composition axis, with relatively balanced contributions from multiple composition variables. The contribution analysis reveals that fat percentage (18%), protein percentage (17%), other solids percentage (16%), lactose percentage (15%), and total solids percentage (15%) all contribute substantially and approximately equally to this component. Somatic cell count contributes minimally (\<2%) to Dim1, indicating it varies independently from the primary milk composition gradient.

Dim2 is primarily driven by milk production quantity, with milk weight (mlk_wt_lbs) contributing 35% of the variance in this component. This represents a milk yield dimension that captures variation in production volume. Other variables show much lower contributions to Dim2, with fat percentage contributing approximately 12% and SCC contributing approximately 10-12%.

```{r}
corr_circle <- function(
    model,
    axes = c(1, 2),
    title = "",
    legend.title = "Contribution",
    gradient.cols = c("#08519C", "#6BAED6", "#FD8D3C")
    ) {
  p <- model %>%
    fviz_pca_var(
      axes = axes,
      col.var = "contrib",
      gradient.cols = gradient.cols,
      repel = TRUE,
      title = title,
      legend.title = legend.title,
      col.circle = "black"
    ) +
    xlim(-1.01, 1.01) +
    ylim(-1.01, 1.01) +
    theme_light() +
    theme(panel.grid = element_blank())
  
  return(p)
}
```

```{r fig.height=4, fig.width=10}
corr_circle(pca_model, title = "Correlation Circle of Variable Relationships") |
  corr_circle(pca_model, axes = c(2, 3))
```

The correlation circle and contribution analysis reveal a three-dimensional structure in the dataset:

-   **Composition Quality Dimension (Dim1):** All major milk composition variables (fat, protein, lactose, other solids, total solids) cluster together with similar contribution patterns, indicating strong positive correlations and suggesting they represent facets of overall milk nutritional quality.
-   **Production Quantity Dimension (Dim2):** Milk weight operates as the primary driver of the second dimension, representing production volume.
-   **Health-Production Trade-off Dimension (Dim2):** The negative correlation between milk weight and somatic cell count along Dim2-axis reveals that higher milk production is associated with lower SCC, while lower production coincides with higher SCC (potentially indicating health issues affecting yield).

```{r}
biplot <- function(model, axes = c(1, 2), col.ind = NULL, legend.title = NULL) {
  p <- model %>%
    fviz_pca_biplot(
      axes = axes,
      geom = "point",
      col.ind = col.ind,
      col.var = "black",
      gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
      col.circle = "white",
      repel = TRUE,
      title = "",
      legend.title = legend.title
    ) +
    theme_light() +
    theme(panel.grid = element_blank())
  
  return(p)
}
```

```{r}
(biplot(pca_model, c(1, 2), lab_data$scc, "SCC\n(10³ cells/mL)") |
  biplot(pca_model, c(2, 3), lab_data$scc, "SCC\n(10³ cells/mL)")
) + plot_layout(guides = "collect")
```

```{r}
(biplot(pca_model, c(1, 3), lab_data$protein, "Protein\n(wt%)") |
  biplot(pca_model, c(3, 4), lab_data$protein, "Protein\n(wt%)")
) + plot_layout(guides = "collect")
```

```{r}
(biplot(pca_model, c(2, 3), lab_data$fat, "Fat\n(wt%)") |
  biplot(pca_model, c(3, 4), lab_data$fat, "Fat\n(wt%)")
) + plot_layout(guides = "collect")
```

```{r}
(biplot(pca_model, c(1, 3), lab_data$lactose, "Lactose\n(wt%)") |
  biplot(pca_model, c(2, 4), lab_data$lactose, "Lactose\n(wt%)")
) + plot_layout(guides = "collect")
```

As previously observed, there's a distinct point that's completely isolated from the main data distribution. This observation indicated by the teal coloring occupies an extreme position in the principal component space, far removed from all other samples. The positioning of these observations as extreme outliers in the PCA space provides compelling evidence that they do not represent genuine biological variation in milk composition. Since principal components are constructed as linear combinations of all input variables, samples containing multiple zero values become mathematically projected to peripheral regions of the PC subspace, where they exert disproportionate influence on the overall data structure.

```{r}
pca_mod <- modeling_data %>%
  select(starts_with("X")) %>%
  PCA(scale.unit = FALSE, graph = FALSE)
```

```{r}
pca_scores <- pca_mod %>%
  pluck("ind", "coord") %>%
  as_tibble() %>%
  bind_cols(modeling_data %>% select(sensor, all_of(targets)), .) %>%
  print()
```

```{r}
ellipse1 <- pca_scores %>% select(starts_with("Dim")) %>% ellipseCoord(pcx = 1, pcy = 2)
```

```{r}
t1 <- round(as.numeric(pca_mod$eig[1,2]), 2)
t2 <- round(as.numeric(pca_mod$eig[2,2]), 2)
t3 <- round(as.numeric(pca_mod$eig[3,2]), 2)
```

```{r}
p1 <- ggplot() +
  geom_polygon(data = ellipse1, aes(x, y), color = "black", fill = "white") +
  geom_point(data = pca_scores, aes(x = Dim.1, y = Dim.2, fill = sensor), shape = 21, size = 3, color = "black") +
  scale_fill_viridis_d(option = "viridis") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  labs(x = glue::glue("Dim1 [{t1}%]"), y = glue::glue("Dim2 [{t2}%]")) +
  theme_grey() +
  theme(
    legend.position = "none",
    aspect.ratio = .7,
    panel.grid = element_blank(),
    panel.background = element_rect(
    colour = "black",
    linewidth = .3
    )
  )

p2 <- ggplot() +
  geom_polygon(data = ellipse1, aes(x, y), color = "black", fill = "white") +
  geom_point(data = pca_scores, aes(x = Dim.1, y = Dim.3, fill = sensor), shape = 21, size = 3, color = "black") +
  scale_fill_viridis_d(option = "viridis") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  labs(
    x = glue::glue("Dim1 [{t1}%]"), 
    y = glue::glue("Dim3 [{t3}%]")
    ) +
  theme_grey() +
   theme(
    aspect.ratio = .7,
    panel.grid = element_blank(),
    panel.background = element_rect(
    colour = "black",
    linewidth = .3
    )
  )
```

```{r fig.height=7, fig.width=8}
wrap_plots(p1, p2, ncol = 1) + 
  plot_layout(guides = "collect")
```

The sensor differences are primarily captured by Dim3 (5.96% variance), while Dim1 and Dim2 (which together explain 71.38% + 14.13% = \~85.4% of the variation) show much more overlap between sensors. This means that the major sources of variation (Dim1 & Dim2) are likely driven by actual sample properties (milk composition differences), while the sensor-to-sensor differences are relegated to a minor component (Dim3). Consequently, models built on this data should be reasonably robust across sensors since the main variation is sample-driven.

```{r}
pca_sensor1 <- modeling_data %>%
  filter(sensor == "sensor_1") %>%
  select(starts_with("X")) %>%
  PCA(scale.unit = FALSE, graph = FALSE)
```

```{r}
scores_sensor1 <- pca_sensor1 %>%
  pluck("ind", "coord") %>%
  as_tibble() %>%
  bind_cols(modeling_data %>% filter(sensor == "sensor_1") %>% select(sensor, all_of(targets)), .) %>%
  print()
```

```{r}
ellipse1 <- scores_sensor1 %>% select(starts_with("Dim")) %>% ellipseCoord(pcx = 1, pcy = 2)
t1 <- round(as.numeric(pca_sensor1$eig[1,2]), 2)
t2 <- round(as.numeric(pca_sensor1$eig[2,2]), 2)
t3 <- round(as.numeric(pca_sensor1$eig[3,2]), 2)
```

```{r}
p1 <- ggplot() +
  geom_polygon(data = ellipse1, aes(x, y), color = "black", fill = "white") +
  geom_point(data = scores_sensor1, aes(x = Dim.1, y = Dim.2, fill = protein), shape = 21, size = 3, color = "black") +
  scale_fill_viridis_c(option = "viridis") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  labs(x = glue::glue("Dim1 [{t1}%]"), y = glue::glue("Dim2 [{t2}%]")) +
  theme_grey() +
  theme(
    legend.position = "none",
    aspect.ratio = .7,
    panel.grid = element_blank(),
    panel.background = element_rect(
    colour = "black",
    linewidth = .3
    )
  )

p2 <- ggplot() +
  geom_polygon(data = ellipse1, aes(x, y), color = "black", fill = "white") +
  geom_point(data = scores_sensor1, aes(x = Dim.1, y = Dim.3, fill = protein), shape = 21, size = 3, color = "black") +
  scale_fill_viridis_c(option = "viridis") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  labs(
    x = glue::glue("Dim1 [{t1}%]"), 
    y = glue::glue("Dim3 [{t3}%]")
    ) +
  theme_grey() +
   theme(
    aspect.ratio = .7,
    panel.grid = element_blank(),
    panel.background = element_rect(
    colour = "black",
    linewidth = .3
    )
  )
```

```{r fig.height=7, fig.width=8}
wrap_plots(p1, p2, ncol = 1) + 
  plot_layout(guides = "collect")
```

# Modeling

## Data splits and preprocessing

```{r}
models_workflow <- function(modeling_data, target_variable) {
  cat(sprintf("\n=== Creating workflow for %s ===\n", target_variable))
  
  data_split <- initial_split(
    modeling_data, 
    prop = 0.8, 
    strata = all_of(target_variable)
    )
  
  train_data <- training(data_split)
  test_data <- testing(data_split)
  
  cat("Training set:", nrow(train_data), "samples\n")
  cat("Test set:", nrow(test_data), "samples\n")
  
  cv_folds <- vfold_cv(train_data, v = 5, strata = all_of(target_variable))
  
  recipe_spec <- recipe(
    as.formula(paste(target_variable, "~ .")), 
    data = train_data
    ) %>%
    step_rm(tube_number, sensor) %>%
    step_zv(all_predictors()) %>%
    step_center(all_numeric_predictors())
  
  return(list(
    split = data_split,
    train = train_data,
    test = test_data,
    cv_folds = cv_folds,
    recipe = recipe_spec
  ))
}
```

## Models tuning and evaluation

```{r}
models_def <- function() {
  cat("Defining model specifications...\n")
  
  rf_spec <- rand_forest() %>%
    set_args(
      mtry = tune(),
      trees = tune(),
      min_n = tune()
      ) %>%
    set_engine("ranger", importance = "impurity") %>%
    set_mode("regression")
  
  elastic_spec <- linear_reg() %>%
    set_args(
      penalty = tune(),
      mixture = tune()
      ) %>%
    set_engine("glmnet") %>%
    set_mode("regression")
  
  xgb_spec <- boost_tree() %>%
    set_args(
      trees = tune(), 
      tree_depth = tune(), 
      learn_rate = tune(), 
      min_n = tune()
      ) %>%
    set_engine("xgboost") %>%
    set_mode("regression")
  
  spls_spec <- pls() %>%
  set_args(
    predictor_prop = tune(), 
    num_comp = tune()
    ) %>%
  set_engine("mixOmics") %>%
  set_mode("regression")
  
  return(list(
    random_forest = rf_spec,
    elastic_net = elastic_spec,
    xgboost = xgb_spec,
    spls = spls_spec
  ))
}
```

```{r}
models_eval <- function(workflow_data, models, target_variable) {
  cat(sprintf("Tuning models for %s...\n", target_variable))
  
  if (Sys.info()["sysname"] == "Darwin") {
    plan(multicore, workers = min(parallel::detectCores() - 2, 8))
    } else {
      plan(multisession, workers = parallel::detectCores() - 1)
      }
  
  results <- list()

  for (model_name in names(models)) {
    cat(sprintf("Processing %s...\n", model_name))

    workflow_spec <- workflow() %>%
      add_recipe(workflow_data$recipe) %>%
      add_model(models[[model_name]])

    if (model_name == "random_forest") {
      tune_grid <- grid_regular(
        mtry(range = c(5, 20)),
        trees(range = c(100, 500)),
        min_n(range = c(5, 20)),
        levels = 3
      )
    } else if (model_name == "elastic_net") {
      tune_grid <- grid_regular(
        penalty(),
        mixture(),
        levels = 5
      )
    } else if (model_name == "xgboost") {
      tune_grid <- grid_regular(
        trees(range = c(100, 500)),
        tree_depth(range = c(3, 10)),
        learn_rate(range = c(0.01, 0.3)),
        min_n(range = c(5, 20)),
        levels = 3
      )
    } else if (model_name == "spls") {
      tune_grid <- grid_regular(
        spectral_features(),
        num_comp(),
        levels = 5
      )
    }

    ctrl_grid <- control_grid(
      verbose = TRUE,
      allow_par = TRUE,
      save_pred = TRUE,
      save_workflow = FALSE,
      parallel_over = "resamples"
    )
    
    tune_results <- workflow_spec %>%
      tune_grid(
        resamples = workflow_data$cv_folds,
        grid = tune_grid,
        metrics = metric_set(rmse, rsq, mae),
        control = ctrl_grid
      )

    best_params <- tune_results %>%
      select_best(metric = "rmse")

    final_workflow <- workflow_spec %>%
      finalize_workflow(best_params)

    final_fit <- final_workflow %>%
      fit(workflow_data$train)

    test_predictions <- final_fit %>%
      predict(workflow_data$test) %>%
      bind_cols(workflow_data$test %>% select(all_of(target_variable)))

    test_metrics <- test_predictions %>%
      metrics(truth = !!sym(target_variable), estimate = .pred)

    results[[model_name]] <- list(
      tune_results = tune_results,
      best_params = best_params,
      final_fit = final_fit,
      test_predictions = test_predictions,
      test_metrics = test_metrics
    )
  }
  return(results)
}
```

## Model performance comparison

```{r}
models_performance <- function(model_results, target_variable) {
  cat(sprintf("\n=== Model Performance Comparison for %s ===\n", target_variable))
  
  performance_summary <- map_dfr(model_results, ~ {
    .x$test_metrics %>%
      filter(.metric %in% c("rmse", "rsq", "mae")) %>%
      select(.metric, .estimate)
  }, .id = "model")

  performance_table <- performance_summary %>%
    pivot_wider(names_from = .metric, values_from = .estimate) %>%
    arrange(rmse)
  
  print(performance_table)
  
  p1 <- performance_summary %>%
    filter(.metric == "rmse") %>%
    ggplot(aes(x = reorder(model, .estimate), y = .estimate, fill = model)) +
    geom_col() +
    coord_flip() +
    labs(
      title = paste("RMSE Comparison -", target_variable),
      x = "Model",
      y = "RMSE",
      fill = "Model"
    ) +
    theme_light() +
    theme(legend.position = "none")
  
  p2 <- performance_summary %>%
    filter(.metric == "rsq") %>%
    ggplot(aes(x = reorder(model, -.estimate), y = .estimate, fill = model)) +
    geom_col() +
    coord_flip() +
    labs(
      title = paste("R² Comparison -", target_variable),
      x = "Model",
      y = "R²",
      fill = "Model"
    ) +
    theme_light() +
    theme(legend.position = "none")
  
  print(p1 / p2)
  return(performance_table)
}
```

## Prediction plots

```{r}
prediction_plots <- function(model_results, target_variable) {
  plots <- map(names(model_results), ~ {
    predictions <- model_results[[.x]]$test_predictions
    
    predictions %>%
      ggplot(aes(x = !!sym(target_variable), y = .pred)) +
      geom_point(alpha = 0.6) +
      geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
      labs(
        title = paste(.x, "-", target_variable),
        x = paste("Actual", target_variable),
        y = paste("Predicted", target_variable)
      ) +
      theme_light()
  })
  
  names(plots) <- names(model_results)
  combined_plot <- wrap_plots(plots, ncol = 2)
  
  print(combined_plot)
  return(plots)
}
```

## Feature Importance

```{r}
feature_importance <- function(model_results, target_variable, top_n = 15) {
  cat(sprintf("\nExtracting feature importance for %s...\n", target_variable))
  
  importance_plots <- list()
  
  for (model_name in names(model_results)) {
    tryCatch({
      if (model_name %in% c("random_forest", "xgboost")) {
        importance_plot <- model_results[[model_name]]$final_fit %>%
          extract_fit_parsnip() %>%
          vip(num_features = top_n) +
          labs(title = paste("Feature Importance -", model_name, "-", target_variable)) +
          theme_minimal()
        importance_plots[[model_name]] <- importance_plot
      }
    }, error = function(e) {
      cat(sprintf("Could not extract importance for %s: %s\n", model_name, e$message))
    })
  }
  
  if (length(importance_plots) > 0) {
    combined_importance <- wrap_plots(importance_plots, ncol = 2)
    print(combined_importance)
  }
  
  return(importance_plots)
}
```

```{r}
run_milk_quality_prediction <- function(
    data, 
    targets = c("fat_percent", "protein_percent", "scc_thous_per_ml", "lactose_percent")
    ) {
  
  cat("=== NIRS Milk Quality Prediction Analysis ===\n")
  
  models <- models_def()
  all_results <- list()
  all_performance <- list()
  
  for (target in targets) {
    cat(sprintf("\n\n### PROCESSING TARGET: %s ###\n", target))
    
    workflow_data <- models_workflow(modeling_data, target)
    model_results <- models_eval(workflow_data, models, target)
    performance <- models_performance(model_results, target)
    pred_plots <- prediction_plots(model_results, target)
    importance_plots <- feature_importance(model_results, target)
    
    all_results[[target]] <- model_results
    all_performance[[target]] <- performance
  }
  cat("\n=== Analysis Complete! ===\n")
  cat("Results stored in all_results and all_performance\n")
  
  return(list(
    results = all_results,
    performance = all_performance,
    modeling_data = modeling_data
  ))
}
```

```{r eval=FALSE, include=FALSE}
cat("Starting NIRS milk quality prediction analysis...\n")
prediction_results <- run_milk_quality_prediction(data = modeling_data)

cat("\n=== SUMMARY: Best Model for Each Target ===\n")

for (target in names(prediction_results$performance)) {
  best_model <- prediction_results$performance[[target]] %>%
    slice_head(n = 1) %>%
    pull(model)
  
  best_rmse <- prediction_results$performance[[target]] %>%
    slice_head(n = 1) %>%
    pull(rmse)
  
  best_rsq <- prediction_results$performance[[target]] %>%
    slice_head(n = 1) %>%
    pull(rsq)
  
  cat(sprintf("%s: %s (RMSE: %.4f, R²: %.4f)\n", target, best_model, best_rmse, best_rsq))
}
```
