---
output: 
  github_document:
    toc: true
    toc_depth: 4
    fig_width: 8
    fig_height: 6
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.retina = 2,
  fig.align = "center",
  dpi = 300,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  cache = FALSE,
  dev = "png"
)
```

# Milk Quality Spectroscopy {.unnumbered}

<!-- badges: start -->

<!-- badges: end -->

```{r}
suppressPackageStartupMessages({
  library(HotellingEllipse)
  library(ConfidenceEllipse)
  library(tidyverse)
  library(magrittr)
  library(reticulate)
  library(viridis)
  library(arrow)
  library(tidymodels)
  library(future)
  library(corrplot)
  library(vip)
  library(patchwork)
  library(yardstick)
  library(skimr)
  library(GGally)
  library(FactoMineR)
  library(factoextra)
  library(gridExtra)
  library(RColorBrewer)
  library(knitr)
  library(naniar)
  library(janitor)
  library(ggforce)
  library(ranger)
  library(glmnet)
  library(xgboost)
  library(mixOmics)
})
```

```{r}
tidymodels_prefer()
theme_set(theme_light(base_size = 12))
```

# Dataset

The data is structured by sensor/stall with each sensor directory containing:

-   **Lab results** (CSV format with tube_no as primary key)
-   **Sensor configuration** (wavelengths, calibration coefficients)
-   **Spectral measurements** (Parquet files, one per milk sample)
-   **Dark spectra** (reference measurements)

## Key data components

**Lab Results**: Milk analysis including fat percentage, protein, somatic cell count and, lactose linked to specific cows and milking sessions.

**Spectral Data**: Raw 16-bit spectral arrays captured during milking, with metadata like temperature, LED current, and integration time. Each spectrum is classified as "dark," "sample," or "empty."

**Sensor Info**: Each sensor has unique wavelength calibrations and measurement parameters that aren't standardized across sensors.

## Important limitations

-   Temperature and LED measurements are raw ADC values, not standardized between sensors
-   Each sensor measures different wavelengths
-   Some timing discrepancies remain due to clock source differences

## Data evolution

The dataset has evolved from initial CSV format to Parquet compression, with spectral data consolidated into array columns rather than individual wavelength columns for more efficient storage and processing.

# Data processing

```{r}
set.seed(123)
```

```{r}
dataset_path <- "Dataset/"
functions_path <- "~/Documents/GitHub/milk-quality-spectroscopy/R/"
```

```{r}
source(paste0(functions_path, "read_sensor_csvs.R"))
source(paste0(functions_path, "read_sensor_parquets.R"))
source(paste0(functions_path, "subtract_dark_spectrum.R"))
source(paste0(functions_path, "process_nir_data.R"))
```

```{r message=FALSE, warning=FALSE}
nir_data <- process_nir_data(verbose = FALSE)
```

```{r}
sensor_wavelengths_list <- nir_data$sensor_data %>%
  select(sensor, wavelengths_nm) %>%
  mutate(
    wavelength_chars = 
      str_remove_all(wavelengths_nm, "\\[|\\]") %>%
      str_split(",\\s*")
    ) %>%
  select(sensor, wavelength_chars) %>%
  deframe()
glimpse(sensor_wavelengths_list)
```

```{r}
spec_data <- nir_data$corrected_spectra %>%
  filter(spec_type == "SAMPLE") %>%
  select(tube_number, sensor, spec_array) %>%
  modify_at("tube_number", as_factor) %>%
  modify_at("sensor", as_factor) %>%
  mutate(
    spec_valid = case_when(
      map_lgl(spec_array, is.null) ~ FALSE,
      map_int(spec_array, length) == 0 ~ FALSE,
      map_lgl(spec_array, ~ all(is.na(.x))) ~ FALSE,
      TRUE ~ TRUE
    )
  ) %>%
  filter(spec_valid) %>%
  rowwise() %>%
  mutate(
    df = list({
      tryCatch(
        {
          values <- as.numeric(unlist(spec_array))
          valid_values <- values[!is.na(values)]
          if (length(valid_values) > 0) {
            setNames(valid_values, paste0("X", seq_along(valid_values)))
          } else {
            NULL
          }
        },
        error = function(e) {
          NULL
        }
      )
    })
  ) %>%
  ungroup() %>%
  filter(!map_lgl(df, is.null)) %>%
  select(-spec_array, -spec_valid) %>%
  unnest_wider(df) %>%
  relocate(sensor, .before = tube_number)
print(spec_data)
```

```{r}
lab_data <- nir_data$lab_results %>%
  select(
    sensor,
    tube_number,
    fat_percent,
    protein_percent,
    scc_thous_per_ml,
    lactose_percent,
    milk_wt_lbs,
    other_solids_percent,
    solids_not_fat_percent,
    total_solids_percent
  ) %>%
  modify_at("tube_number", as_factor) %>%
  modify_at("sensor", as_factor) %>%
  filter(
    !is.na(fat_percent),
    !is.na(protein_percent),
    !is.na(scc_thous_per_ml),
    !is.na(lactose_percent)
  ) %>%
  rename(
    fat = fat_percent,
    protein = protein_percent,
    scc = scc_thous_per_ml,
    lactose = lactose_percent,
    weight = milk_wt_lbs,
    misc_solids = other_solids_percent,
    nonfat_solids = solids_not_fat_percent,
    tot_solids = total_solids_percent
  ) %>%
  print()
```

```{r}
targets <- lab_data %>% select(fat, protein, scc, lactose) %>% names()
meta <- lab_data %>% select(-sensor, -tube_number, -all_of(targets)) %>% names()
```

```{r}
modeling_data <- spec_data %>%
  left_join(lab_data, by = c("sensor", "tube_number")) %>%
  filter(if_all(starts_with("X_"), ~ !is.na(.x))) %>%
  relocate(sensor, .before = tube_number) %>%
  relocate(all_of(targets), .after = tube_number) %>%
  relocate(all_of(meta), .after = all_of(targets))
```

```{r}
cat("Final dataset dimensions:", nrow(modeling_data), "x", ncol(modeling_data), "\n")
cat("Number of wavelenghts:", sum(grepl("^X", names(modeling_data))), "\n")
```

```{r}
modeling_data %>%
  select(sensor, starts_with("X")) %>%
  group_by(sensor) %>%
  summarise(
    observations = n(),
    variables = ncol(.) - 1,
    total_cells = n() * ncol(.),
    missing_values = sum(is.na(.)),
    completeness = round((1 - sum(is.na(.)) / (n() * ncol(.))) * 100, 1),
    .groups = 'drop'
  ) %>%
  arrange(sensor)
```

## Observations by Sensor

```{r}
sensor_summary <- modeling_data %>%
    select(sensor, tube_number, starts_with("X")) %>%
    group_by(sensor, tube_number) %>%
    summarise(
        observations = n(),
        variables = ncol(.) - 2,
        .groups = 'drop'
    )
```

```{r}
sensor_summary %>%
    group_by(sensor) %>%
    summarise(total_obs = sum(observations), .groups = 'drop') %>%
    mutate(sensor_num = as.numeric(str_extract(sensor, "\\d+"))) %>%
    arrange(sensor_num) %>%
    ggplot(aes(x = reorder(sensor, sensor_num), y = total_obs)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_text(
      aes(label = scales::comma(total_obs)),
      vjust = -0.3, 
      size = 3.5, 
      fontface = "bold"
      ) +
    labs(
        title = "Total Observations by Sensor",
        subtitle = "Comparison across all 12 sensors",
        x = "",
        y = "Total Observations"
    ) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank()
    ) +
    scale_y_continuous(labels = scales::comma)
```

```{r message=FALSE, warning=FALSE}
sensor_summary %>%
  mutate(sensor_num = as.numeric(str_extract(sensor, "\\d+"))) %>%
  group_by(sensor) %>%
  arrange(tube_number) %>%
  mutate(
    tube_position = row_number(),
    max_tubes = max(tube_position)
  ) %>%
  ungroup() %>%
  ggplot(
    aes(
      x = tube_position, 
      y = reorder(sensor, -sensor_num), 
      fill = observations
      )
    ) +
  geom_tile(color = "white", size = 0.3) +
  scale_fill_gradient2(
    low = "lightblue", mid = "yellow", high = "red",
    midpoint = median(sensor_summary$observations),
    name = "Observations"
  ) +
  labs(
    title = "Observations Heatmap: All Sensors",
    subtitle = "X-axis shows tube position (1st, 2nd, 3rd, etc.) within each sensor",
    x = "Tube Position (ordered within sensor)",
    y = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 8)
  ) +
  scale_x_continuous(
    breaks = seq(
      1, 
      max(
        sensor_summary %>% 
          group_by(sensor) %>% 
          summarise(n = n()) %>% 
          pull(n)
        ), 
      by = 5
      )
    )
```

## Measurement timeline

```{r}
temporal_summary <- nir_data$corrected_spectra %>%
  mutate(datetime = as.POSIXct(datetime)) %>%
  group_by(sensor) %>%
  filter(spec_type == "SAMPLE") %>%
  drop_na() %>%
  summarise(
    min_time = min(datetime, na.rm = TRUE),
    max_time = max(datetime, na.rm = TRUE),
    time_span_hours = as.numeric(difftime(
      max(datetime, na.rm = TRUE),
      min(datetime, na.rm = TRUE),
      units = "hours"
    )),
    .groups = "drop"
  ) %>%
  print()
```

```{r}
nir_data$corrected_spectra %>%
  mutate(datetime = as.POSIXct(datetime)) %>%
  filter(!is.na(datetime)) %>%
  filter(spec_type == "SAMPLE") %>%
  drop_na() %>%
  ggplot(aes(x = datetime, y = sensor, color = sensor)) +
  geom_point(
    alpha = 0.6,
    size = 0.5,
    position = position_jitter(height = 0.2, width = 0)
  ) +
  ggsci::scale_color_ucscgb() +
  labs(
    title = "Measurement Timeline Across All Sensors",
    subtitle = sprintf(
      "Total measurements: %d | Date range: %s to %s",
      nrow(nir_data$corrected_spectra %>% filter(spec_type == "SAMPLE")),
      format(min(temporal_summary$min_time), "%Y-%m-%d"),
      format(max(temporal_summary$max_time), "%Y-%m-%d")
    ),
    x = "Date/Time",
    y = "",
    caption = "Each point represents one spectral measurement"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "gray40")
  )
```

## Spectra

```{r}
source(paste0(functions_path, "visualize_nirs_time_series.R"))
```

```{r message=FALSE, warning=FALSE}
visualize_nirs_time_series(nir_data, "sensor_1", min_measurements = 2, seed = 100)
```

# Exploratory Data Analysis

## Descriptive statistics for milk quality target variables

```{r}
source(paste0(functions_path, "calculate_target_statistics.R"))
source(paste0(functions_path, "create_distribution_plots.R"))
source(paste0(functions_path, "analyze_correlations.R"))
source(paste0(functions_path, "analyze_by_groups.R"))
source(paste0(functions_path, "identify_outliers.R"))
```

```{r}
run_descriptive_analysis <- function(nir_data) {
  cat("Starting comprehensive descriptive analysis of target variables...\n\n")
  stats_results <- calculate_target_statistics(nir_data)
  dist_plots <- create_distribution_plots(stats_results$targets_data)
  correlations <- analyze_correlations(stats_results$targets_data)
  group_analysis <- analyze_by_groups(stats_results$targets_data)
  outliers <- identify_outliers(stats_results$targets_data)
  
  cat("=== Comprehensive Data Summary (skimr) ===\n")
  target_vars <- c("fat_percent", "protein_percent", "scc_thous_per_ml", "lactose_percent")
  skim_summary <- stats_results$targets_data %>%
    select(all_of(target_vars)) %>%
    skim()
  
  print(skim_summary)
  
  cat("\n=== Analysis Complete ===\n")
  cat("Key Insights:\n")
  cat("- Total samples analyzed:", nrow(stats_results$targets_data), "\n")
  cat("- Variables analyzed:", length(target_vars), "\n")
  cat("- Correlation range:", round(min(correlations[correlations != 1]), 3), 
      "to", round(max(correlations[correlations != 1]), 3), "\n")
  
  return(list(
    summary_stats = stats_results$summary_stats,
    targets_data = stats_results$targets_data,
    distributions = dist_plots,
    correlations = correlations,
    group_analysis = group_analysis,
    outliers = outliers,
    skim_summary = skim_summary
  ))
}
```

```{r message=FALSE, warning=FALSE}
descriptive_results <- run_descriptive_analysis(nir_data)
cat("\n=== Quick Summary ===\n")
cat("Mean values:\n")
descriptive_results$summary_stats %>%
  select(variable, mean) %>%
  mutate(mean = round(mean, 3)) %>%
  print()

cat("\nCoefficient of Variation (CV%):\n")
descriptive_results$summary_stats %>%
  select(variable, cv) %>%
  mutate(cv = round(cv, 2)) %>%
  arrange(desc(cv)) %>%
  print()
```

## PCA analysis of milk quality targets

A principal component analysis (PCA) was conducted on four key milk composition variables: fat percentage, protein percentage, somatic cell count (SCC) per milliliter (thousands), and lactose percentage. The analysis aimed to understand the underlying structure of relationships between these variables and assess potential multicollinearity for subsequent modeling efforts.

```{r}
pca_data <- nir_data$lab_results %>%
    select(
      fat_percent, 
      protein_percent, 
      scc_thous_per_ml, 
      lactose_percent,
      dim_days, # days since the cow started producing milk
      milk_wt_lbs, # weight of the milk in pounds
      other_solids_percent, # other solids percentage in the milk
      solids_not_fat_percent, # total solids that are not fat as a percentage
      total_solids_percent, # total solids as a percentage in the milk
      mun_mg_per_dl # total solids in milligrams per deciliter
      ) %>%
    filter(
      !is.na(fat_percent), 
      !is.na(protein_percent), 
      !is.na(scc_thous_per_ml), 
      !is.na(lactose_percent)
    )
```

```{r}
quanti_vars <- c(
    "fat_percent", "protein_percent", "scc_thous_per_ml", "lactose_percent",
    "other_solids_percent", "solids_not_fat_percent", "total_solids_percent",
    "mun_mg_per_dl", "milk_wt_lbs", "dim_days"
    )
```

```{r}
quanti_indices <- which(names(pca_data) %in% quanti_vars)
```

```{r}
pca_result <- PCA(
    pca_data,
    scale.unit = TRUE,
    quanti.sup = if(length(setdiff(quanti_indices, 1:10)) > 0) {
      setdiff(quanti_indices, 1:10)
    } else NULL,
    graph = FALSE
  )
```

```{r}
scree_plot <- fviz_eig(
    pca_result, 
    addlabels = TRUE, 
    ylim = c(0, 60),
    title = "Scree Plot",
    xlab = "Principal Components",
    ylab = "Percentage of Variance Explained"
    ) +
    theme_light() +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red", alpha = 0.7) +
    labs(caption = "Red line indicates 10% variance threshold")

eigenvalues <- get_eigenvalue(pca_result)
cum_var_plot <- as.data.frame(eigenvalues) %>%
    mutate(component = 1:nrow(.)) %>%
    ggplot(aes(x = component, y = cumulative.variance.percent)) +
    geom_line(size = 1.2, color = "steelblue") +
    geom_point(size = 3, color = "steelblue") +
    geom_hline(yintercept = 80, linetype = "dashed", color = "red", alpha = 0.7) +
    scale_x_continuous(breaks = 1:nrow(eigenvalues)) +
    labs(
      title = "Cumulative Variance Explained",
      x = "Principal Components",
      y = "Cumulative Variance (%)",
      caption = "Red line indicates 80% variance threshold"
    ) +
    theme_light()
```

```{r}
scree_plot | cum_var_plot
```

The PCA extracted two principal components that collectively explained 67.6% of the total variance in the dataset. The first principal component accounted for 51.1% of the variance, while the second component explained an additional 16.5%.

```{r}
contrib_pc1 <- fviz_contrib(
  pca_result,
  choice = "var",
  axes = 1,
  top = 10,
  title = "Dim1",
  fill = "steelblue",
  color = "steelblue"
  ) +
  labs(x = "") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

contrib_pc2 <- fviz_contrib(
  pca_result,
  choice = "var",
  axes = 2,
  top = 10,
  title = "Dim2",
  fill = "coral",
  color = "coral"
  ) +
  labs(x = "") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

contrib_combined <- fviz_contrib(
  pca_result,
  choice = "var",
  axes = 1:2,
  title = "Dim1 and Dim2",
  fill = "darkgreen",
  color = "darkgreen"
  ) +
  labs(x = "") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
contrib_pc1 | contrib_pc2 | contrib_combined
```

Dim1 represents a milk composition axis, with relatively balanced contributions from multiple composition variables. The contribution analysis reveals that fat percentage (18%), protein percentage (17%), other solids percentage (16%), lactose percentage (15%), and total solids percentage (15%) all contribute substantially and approximately equally to this component. Somatic cell count contributes minimally (\<2%) to Dim1, indicating it varies independently from the primary milk composition gradient.

Dim2 is primarily driven by milk production quantity, with milk weight (mlk_wt_lbs) contributing 35% of the variance in this component. This represents a milk yield dimension that captures variation in production volume. Other variables show much lower contributions to Dim2, with fat percentage contributing approximately 12% and SCC contributing approximately 10-12%.

```{r}
p1 <- fviz_pca_var(
  pca_result,
  col.var = "cos2",
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  col.circle = "white",
  repel = TRUE,
  title = "Quality Assessment of Variable Projections",
  legend.title = "Cos2"
  ) +
  theme_light() +
  theme(panel.grid = element_blank())
```

```{r}
p2 <- fviz_pca_var(
  pca_result,
  col.var = "contrib",
  gradient.cols = c("darkblue", "green", "darkred"),
  repel = TRUE,
  title = "Correlation Circle of Variable Relationships",
  legend.title = "Contribution"
  ) +
  annotate(
    "circle",
    x = 0, y = 0,
    color = "grey70",
    linetype = "solid",
    linewidth = 0.1
  ) +
  xlim(-1.01, 1.01) + 
  ylim(-1.01, 1.01) +
  theme_light() +
  theme(panel.grid = element_blank())
```

```{r fig.height=5, fig.width=10}
p1 | p2
```

The correlation circle and contribution analysis reveal a three-dimensional structure in the dataset:

-   **Composition Quality Dimension (Dim1):** All major milk composition variables (fat, protein, lactose, other solids, total solids) cluster together with similar contribution patterns, indicating strong positive correlations and suggesting they represent facets of overall milk nutritional quality.
-   **Production Quantity Dimension (Dim2):** Milk weight operates as the primary driver of the second dimension, representing production volume.
-   **Health-Production Trade-off Dimension (Dim2):** The negative correlation between milk weight and somatic cell count along Dim2-axis reveals that higher milk production is associated with lower SCC, while lower production coincides with higher SCC (potentially indicating health issues affecting yield).

```{r}
pca_mod <- modeling_data %>%
  select(starts_with("X")) %>%
  PCA(scale.unit = FALSE, graph = FALSE)
```

```{r}
pca_scores <- pca_mod %>%
  pluck("ind", "coord") %>%
  as_tibble() %>%
  bind_cols(modeling_data %>% select(sensor, all_of(targets)), .) %>%
  print()
```

```{r}
ellipse1 <- pca_scores %>% select(starts_with("Dim")) %>% ellipseCoord(pcx = 1, pcy = 2)
```

```{r}
t1 <- round(as.numeric(pca_mod$eig[1,2]), 2)
t2 <- round(as.numeric(pca_mod$eig[2,2]), 2)
t3 <- round(as.numeric(pca_mod$eig[3,2]), 2)
```

```{r}
p1 <- ggplot() +
  geom_polygon(data = ellipse1, aes(x, y), color = "black", fill = "white") +
  geom_point(data = pca_scores, aes(x = Dim.1, y = Dim.2, fill = sensor), shape = 21, size = 3, color = "black") +
  scale_fill_viridis_d(option = "viridis") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  labs(x = glue::glue("Dim1 [{t1}%]"), y = glue::glue("Dim2 [{t2}%]")) +
  theme_grey() +
  theme(
    legend.position = "none",
    aspect.ratio = .7,
    panel.grid = element_blank(),
    panel.background = element_rect(
    colour = "black",
    linewidth = .3
    )
  )

p2 <- ggplot() +
  geom_polygon(data = ellipse1, aes(x, y), color = "black", fill = "white") +
  geom_point(data = pca_scores, aes(x = Dim.1, y = Dim.3, fill = sensor), shape = 21, size = 3, color = "black") +
  scale_fill_viridis_d(option = "viridis") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  labs(
    x = glue::glue("Dim1 [{t1}%]"), 
    y = glue::glue("Dim3 [{t3}%]")
    ) +
  theme_grey() +
   theme(
    aspect.ratio = .7,
    panel.grid = element_blank(),
    panel.background = element_rect(
    colour = "black",
    linewidth = .3
    )
  )
```

```{r fig.height=7, fig.width=8}
wrap_plots(p1, p2, ncol = 1) + 
  plot_layout(guides = "collect")
```

The sensor differences are primarily captured by Dim3 (5.96% variance), while Dim1 and Dim2 (which together explain 71.38% + 14.13% = \~85.4% of the variation) show much more overlap between sensors. This means that the major sources of variation (Dim1 & Dim2) are likely driven by actual sample properties (milk composition differences), while the sensor-to-sensor differences are relegated to a minor component (Dim3). Consequently, models built on this data should be reasonably robust across sensors since the main variation is sample-driven.

```{r}
pca_sensor1 <- modeling_data %>%
  filter(sensor == "sensor_1") %>%
  select(starts_with("X")) %>%
  PCA(scale.unit = FALSE, graph = FALSE)
```

```{r}
scores_sensor1 <- pca_sensor1 %>%
  pluck("ind", "coord") %>%
  as_tibble() %>%
  bind_cols(modeling_data %>% filter(sensor == "sensor_1") %>% select(sensor, all_of(targets)), .) %>%
  print()
```

```{r}
ellipse1 <- scores_sensor1 %>% select(starts_with("Dim")) %>% ellipseCoord(pcx = 1, pcy = 2)
t1 <- round(as.numeric(pca_sensor1$eig[1,2]), 2)
t2 <- round(as.numeric(pca_sensor1$eig[2,2]), 2)
t3 <- round(as.numeric(pca_sensor1$eig[3,2]), 2)
```

```{r}
p1 <- ggplot() +
  geom_polygon(data = ellipse1, aes(x, y), color = "black", fill = "white") +
  geom_point(data = scores_sensor1, aes(x = Dim.1, y = Dim.2, fill = protein), shape = 21, size = 3, color = "black") +
  scale_fill_viridis_c(option = "viridis") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  labs(x = glue::glue("Dim1 [{t1}%]"), y = glue::glue("Dim2 [{t2}%]")) +
  theme_grey() +
  theme(
    legend.position = "none",
    aspect.ratio = .7,
    panel.grid = element_blank(),
    panel.background = element_rect(
    colour = "black",
    linewidth = .3
    )
  )

p2 <- ggplot() +
  geom_polygon(data = ellipse1, aes(x, y), color = "black", fill = "white") +
  geom_point(data = scores_sensor1, aes(x = Dim.1, y = Dim.3, fill = protein), shape = 21, size = 3, color = "black") +
  scale_fill_viridis_c(option = "viridis") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "black", linewidth = .2) +
  labs(
    x = glue::glue("Dim1 [{t1}%]"), 
    y = glue::glue("Dim3 [{t3}%]")
    ) +
  theme_grey() +
   theme(
    aspect.ratio = .7,
    panel.grid = element_blank(),
    panel.background = element_rect(
    colour = "black",
    linewidth = .3
    )
  )
```

```{r fig.height=7, fig.width=8}
wrap_plots(p1, p2, ncol = 1) + 
  plot_layout(guides = "collect")
```

# Modeling

## Data splits and preprocessing

```{r}
models_workflow <- function(modeling_data, target_variable) {
  cat(sprintf("\n=== Creating workflow for %s ===\n", target_variable))
  
  data_split <- initial_split(
    modeling_data, 
    prop = 0.8, 
    strata = all_of(target_variable)
    )
  
  train_data <- training(data_split)
  test_data <- testing(data_split)
  
  cat("Training set:", nrow(train_data), "samples\n")
  cat("Test set:", nrow(test_data), "samples\n")
  
  cv_folds <- vfold_cv(train_data, v = 5, strata = all_of(target_variable))
  
  recipe_spec <- recipe(
    as.formula(paste(target_variable, "~ .")), 
    data = train_data
    ) %>%
    step_rm(tube_number, sensor) %>%
    step_zv(all_predictors()) %>%
    step_center(all_numeric_predictors())
  
  return(list(
    split = data_split,
    train = train_data,
    test = test_data,
    cv_folds = cv_folds,
    recipe = recipe_spec
  ))
}
```

## Models tuning and evaluation

```{r}
models_def <- function() {
  cat("Defining model specifications...\n")
  
  rf_spec <- rand_forest() %>%
    set_args(
      mtry = tune(),
      trees = tune(),
      min_n = tune()
      ) %>%
    set_engine("ranger", importance = "impurity") %>%
    set_mode("regression")
  
  elastic_spec <- linear_reg() %>%
    set_args(
      penalty = tune(),
      mixture = tune()
      ) %>%
    set_engine("glmnet") %>%
    set_mode("regression")
  
  xgb_spec <- boost_tree() %>%
    set_args(
      trees = tune(), 
      tree_depth = tune(), 
      learn_rate = tune(), 
      min_n = tune()
      ) %>%
    set_engine("xgboost") %>%
    set_mode("regression")
  
  spls_spec <- pls() %>%
  set_args(
    predictor_prop = tune(), 
    num_comp = tune()
    ) %>%
  set_engine("mixOmics") %>%
  set_mode("regression")
  
  return(list(
    random_forest = rf_spec,
    elastic_net = elastic_spec,
    xgboost = xgb_spec,
    spls = spls_spec
  ))
}
```

```{r}
models_eval <- function(workflow_data, models, target_variable) {
  cat(sprintf("Tuning models for %s...\n", target_variable))
  
  if (Sys.info()["sysname"] == "Darwin") {
    plan(multicore, workers = min(parallel::detectCores() - 2, 8))
    } else {
      plan(multisession, workers = parallel::detectCores() - 1)
      }
  
  results <- list()

  for (model_name in names(models)) {
    cat(sprintf("Processing %s...\n", model_name))

    workflow_spec <- workflow() %>%
      add_recipe(workflow_data$recipe) %>%
      add_model(models[[model_name]])

    if (model_name == "random_forest") {
      tune_grid <- grid_regular(
        mtry(range = c(5, 20)),
        trees(range = c(100, 500)),
        min_n(range = c(5, 20)),
        levels = 3
      )
    } else if (model_name == "elastic_net") {
      tune_grid <- grid_regular(
        penalty(),
        mixture(),
        levels = 5
      )
    } else if (model_name == "xgboost") {
      tune_grid <- grid_regular(
        trees(range = c(100, 500)),
        tree_depth(range = c(3, 10)),
        learn_rate(range = c(0.01, 0.3)),
        min_n(range = c(5, 20)),
        levels = 3
      )
    } else if (model_name == "spls") {
      tune_grid <- grid_regular(
        spectral_features(),
        num_comp(),
        levels = 5
      )
    }

    ctrl_grid <- control_grid(
      verbose = TRUE,
      allow_par = TRUE,
      save_pred = TRUE,
      save_workflow = FALSE,
      parallel_over = "resamples"
    )
    
    tune_results <- workflow_spec %>%
      tune_grid(
        resamples = workflow_data$cv_folds,
        grid = tune_grid,
        metrics = metric_set(rmse, rsq, mae),
        control = ctrl_grid
      )

    best_params <- tune_results %>%
      select_best(metric = "rmse")

    final_workflow <- workflow_spec %>%
      finalize_workflow(best_params)

    final_fit <- final_workflow %>%
      fit(workflow_data$train)

    test_predictions <- final_fit %>%
      predict(workflow_data$test) %>%
      bind_cols(workflow_data$test %>% select(all_of(target_variable)))

    test_metrics <- test_predictions %>%
      metrics(truth = !!sym(target_variable), estimate = .pred)

    results[[model_name]] <- list(
      tune_results = tune_results,
      best_params = best_params,
      final_fit = final_fit,
      test_predictions = test_predictions,
      test_metrics = test_metrics
    )
  }
  return(results)
}
```

## Model performance comparison

```{r}
models_performance <- function(model_results, target_variable) {
  cat(sprintf("\n=== Model Performance Comparison for %s ===\n", target_variable))
  
  performance_summary <- map_dfr(model_results, ~ {
    .x$test_metrics %>%
      filter(.metric %in% c("rmse", "rsq", "mae")) %>%
      select(.metric, .estimate)
  }, .id = "model")

  performance_table <- performance_summary %>%
    pivot_wider(names_from = .metric, values_from = .estimate) %>%
    arrange(rmse)
  
  print(performance_table)
  
  p1 <- performance_summary %>%
    filter(.metric == "rmse") %>%
    ggplot(aes(x = reorder(model, .estimate), y = .estimate, fill = model)) +
    geom_col() +
    coord_flip() +
    labs(
      title = paste("RMSE Comparison -", target_variable),
      x = "Model",
      y = "RMSE",
      fill = "Model"
    ) +
    theme_light() +
    theme(legend.position = "none")
  
  p2 <- performance_summary %>%
    filter(.metric == "rsq") %>%
    ggplot(aes(x = reorder(model, -.estimate), y = .estimate, fill = model)) +
    geom_col() +
    coord_flip() +
    labs(
      title = paste("R² Comparison -", target_variable),
      x = "Model",
      y = "R²",
      fill = "Model"
    ) +
    theme_light() +
    theme(legend.position = "none")
  
  print(p1 / p2)
  return(performance_table)
}
```

## Prediction plots

```{r}
prediction_plots <- function(model_results, target_variable) {
  plots <- map(names(model_results), ~ {
    predictions <- model_results[[.x]]$test_predictions
    
    predictions %>%
      ggplot(aes(x = !!sym(target_variable), y = .pred)) +
      geom_point(alpha = 0.6) +
      geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
      labs(
        title = paste(.x, "-", target_variable),
        x = paste("Actual", target_variable),
        y = paste("Predicted", target_variable)
      ) +
      theme_light()
  })
  
  names(plots) <- names(model_results)
  combined_plot <- wrap_plots(plots, ncol = 2)
  
  print(combined_plot)
  return(plots)
}
```

## Feature Importance

```{r}
feature_importance <- function(model_results, target_variable, top_n = 15) {
  cat(sprintf("\nExtracting feature importance for %s...\n", target_variable))
  
  importance_plots <- list()
  
  for (model_name in names(model_results)) {
    tryCatch({
      if (model_name %in% c("random_forest", "xgboost")) {
        importance_plot <- model_results[[model_name]]$final_fit %>%
          extract_fit_parsnip() %>%
          vip(num_features = top_n) +
          labs(title = paste("Feature Importance -", model_name, "-", target_variable)) +
          theme_minimal()
        importance_plots[[model_name]] <- importance_plot
      }
    }, error = function(e) {
      cat(sprintf("Could not extract importance for %s: %s\n", model_name, e$message))
    })
  }
  
  if (length(importance_plots) > 0) {
    combined_importance <- wrap_plots(importance_plots, ncol = 2)
    print(combined_importance)
  }
  
  return(importance_plots)
}
```

```{r}
run_milk_quality_prediction <- function(
    data, 
    targets = c("fat_percent", "protein_percent", "scc_thous_per_ml", "lactose_percent")
    ) {
  
  cat("=== NIRS Milk Quality Prediction Analysis ===\n")
  
  models <- models_def()
  all_results <- list()
  all_performance <- list()
  
  for (target in targets) {
    cat(sprintf("\n\n### PROCESSING TARGET: %s ###\n", target))
    
    workflow_data <- models_workflow(modeling_data, target)
    model_results <- models_eval(workflow_data, models, target)
    performance <- models_performance(model_results, target)
    pred_plots <- prediction_plots(model_results, target)
    importance_plots <- feature_importance(model_results, target)
    
    all_results[[target]] <- model_results
    all_performance[[target]] <- performance
  }
  cat("\n=== Analysis Complete! ===\n")
  cat("Results stored in all_results and all_performance\n")
  
  return(list(
    results = all_results,
    performance = all_performance,
    modeling_data = modeling_data
  ))
}
```

```{r eval=FALSE, include=FALSE}
cat("Starting NIRS milk quality prediction analysis...\n")
prediction_results <- run_milk_quality_prediction(data = modeling_data)

cat("\n=== SUMMARY: Best Model for Each Target ===\n")

for (target in names(prediction_results$performance)) {
  best_model <- prediction_results$performance[[target]] %>%
    slice_head(n = 1) %>%
    pull(model)
  
  best_rmse <- prediction_results$performance[[target]] %>%
    slice_head(n = 1) %>%
    pull(rmse)
  
  best_rsq <- prediction_results$performance[[target]] %>%
    slice_head(n = 1) %>%
    pull(rsq)
  
  cat(sprintf("%s: %s (RMSE: %.4f, R²: %.4f)\n", target, best_model, best_rmse, best_rsq))
}
```
